---
title: "Tidy Data"
author: "Ian O'Keeffe"
date: "2026-09-29"
output: html_document
---
This file is for doing data tidying
```{r}
library(tidyverse)

data('mtcars')
```

## Let's tidy some data
```{r}
pulse_df = 
  haven::read_sas('data/public_pulse_data.sas7bdat') %>% 
  janitor::clean_names()
```
Data is embedded in column names, let's fix that. baseline, one month, 6 months are column names. we'll pivot out and add new names.

names_prefix clears leading characters. names_to gives a new column with old column heads
```{r}
pulse_tidy_df = 
  pulse_df %>% 
  pivot_longer(
    bdi_score_bl:bdi_score_12m,
    names_to = "visit",
    names_prefix = "bdi_score_",
    values_to = "bdi_score"
  ) %>% 
  mutate(
    visit = replace(visit, visit == "bl", "00m")
  )
pulse_tidy_df
```
```{r}
litters = 
  read_csv("data/FAS_litters.csv", na = c("", "NA", ".")) %>% 
  janitor::clean_names()

litters
```

```{r}


litters = 
  read_csv("data/FAS_litters.csv", na = c("", "NA", ".")) %>% 
  janitor::clean_names() %>% 
  select(litter_number, gd0_weight, gd18_weight) %>% 
  pivot_longer(
    gd0_weight:gd18_weight,
    names_to = "gd",
    values_to = "weight"
  ) %>% 
  mutate(
    gd = case_match(
      gd,
      "gd0_weight" ~ 0,
      "gd18_weight" ~ 18,
    )
  )

litters
```
## tidy data
```{r}
analysis_df = 
  tibble(
    groups = c("treatment", "treatment", "placebo", "placebo"),
    time = c("pre", "post", "pre", "post"),
    mean_outcome = c(4.0, 8.0, 3.5, 4.6)
  )
analysis_df
```
## Let's untidy it for readability. knitr::kable is creating a table for our markdown output.
```{r}
analysis_df %>% 
  pivot_wider(
    names_from = time,
    values_from = mean_outcome
  ) %>% 
  knitr::kable()
```
## Bind some rows

first, import each lotr movie table
```{r}
fellowship_df = 
  readxl::read_excel("data/LotR_Words.xlsx", range = "B3:D6") %>% 
  mutate(movie = "fellowship")

towers_df = 
  readxl::read_excel("data/LotR_Words.xlsx", range = "F3:H6") %>% 
  mutate(movie = "towers")

return_df = 
  readxl::read_excel("data/LotR_Words.xlsx", range = "J3:L6") %>% 
  mutate(movie = "return")
```

```{r}

lotr_df = 
  bind_rows(fellowship_df, towers_df, return_df) %>% 
  janitor::clean_names() %>% 
  relocate(movie) %>% 
  pivot_longer(
    female:male,
    names_to = "gender",
    values_to = "words"
  )


lotr_df
```
## join FAS datasets
```{r}
pups_df = 
  read_csv("data/FAS_pups.csv",
           skip = 3,
           na = c("", ".", "NA")
  ) %>% 
  janitor::clean_names()

litters_df = 
  read_csv(
      "data/FAS_litters.csv", 
      na = c("", "NA", ".")
  ) %>% 
  janitor::clean_names()

fas_df = 
  left_join(pups_df, litters_df, by = "litter_number")

```

